Exploratory Analysis

### 2023-11-29 Code/Draft1Rmd.Rmd
Added Line Plots

library(tidyverse)
library(ggplot2)

@@

>>>>

library(tidyverse)
library(ggplot2)
source("Functions.R")

@@

```{r}
#Create individual data sets for Oregon, Washington, and California
or_health_ranking <- health_ranking[health_ranking$state == "Oregon",]
ca_health_ranking <- health_ranking[health_ranking$state == "California",]
wa_health_ranking <- health_ranking[health_ranking$state == "Washington",]

#Now create yearly averages for the states from the healthrank_to_yearly function from Functions.R:
or_yearly <- healthrank_to_yearly(or_health_ranking)
ca_yearly <- healthrank_to_yearly(ca_health_ranking)
wa_yearly <- healthrank_to_yearly(wa_health_ranking)

ggplot()+
  geom_line(data = or_yearly, aes(x = year, y = combined_overdose_deaths, color = "Oregon")) +
  geom_line(data = ca_yearly, aes(x = year, y = combined_overdose_deaths, color = "California")) +
  geom_line(data = wa_yearly, aes(x = year, y = combined_overdose_deaths, color = "Washington")) +
  scale_color_manual(name = "State",
                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
  labs(y = "Overdose Deaths") +
  theme_bw()

#Now do a log transformation to better show percent differences due to population size differences
ggplot()+
  geom_line(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon")) +
  geom_line(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California")) +
  geom_line(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington")) +
  scale_color_manual(name = "State",
                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
  theme_bw()
```

### 2023-11-29 Code/Draft1Rmd.tex
Added Line Plots

---
title: "Exploratory Analysis"
author: "James Santarpio and Mardi Elings"
date: "2023-11-30"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(ggplot2)
source("Functions.R")

url <- c("https://raw.githubusercontent.com/mardimardimardi/ECNS_460_TopicData/main/Data/Cleaned%20Data/CleanedStateHealthRankings/MergedAllCountyRanks.csv")
# Read csv
data <- read.csv(url)

# Remove 2023- 2023 is a duplicate of 2022 for much of the data
health_ranking <- data %>%
  filter(year <= 2023) %>%
  filter(! (county %in% c("year_state_total")))

```
# Data Analytics Term Project
## Project Desription

### Topic:
The effect of drug decriminalization in Oregon on overdose deaths and well-being compared to California and Washington.

### Why it Matters 
The study of drug decriminalization and overdose rates in Oregon has potential for significant policy and societal relevance. There are many debates on the benefits and drawbacks that drug use and decriminalization has on potential for policy implications involving economics, public health, criminal justice, and community well-being.

### Research Question and Motivation
What is the effect of drug decriminalization on overdose deaths and well-being? A policy change like Oregon's can have significant consequences on populations who are most at risk of overdose deaths, economic impacts on health care costs, and overall well-being. Investigating the effects with county level data, we used a  difference-in-difference model with fixed effects to assess the impact of the policy change in Oregon. By comparing the outcomes in Oregon before and after the intervention to a control group of Washington and California counties as control groups.

### Description of Data

#### Overdose Mortality by County in the US:
The Overdose Mortality by County in the US is from the CDC Center for Health Statistics for the years 2020-2023. These provisional county-Level drug Overdose death counts are based on death records and is received by the NCHS receives the death records from the state vital registration offices through the Vital Statistics Cooperative Program (VSCP). The provisional counts are reported by the county were the decedent resided, not necessarily where they died.

#### County Well-Being Rankings:
The County Well-Being Rankings are for the years 2014-2023 and is collected by County Health Rankings & Roadmaps CHR&R, a program of the University of Wisconsin Population Health Institute. The health rankings include deaths which include overdose deaths, years of potential life lost, mental health statistics, income, and other health public health rankings.

#### Overdose by Drug:
The Overdose by drug dataset contains information on non-fatal overdose data by state for 2018-2023. Provided by the CDC, DOSE (Drug Overdose Surveillance and Epidemiology) collects health department reports in each state.

@@

#### How are they related? 
The Well-Being Ranks, Overdose Mortality by County contain county fips codes, county names and year in relation. The Overdose by Drug has a relationship of total non-fatal drug overdoses in each state.

## Data Processing 
For the county health rankings data, we originally had years 2014-2023, but after further inspection we found the way overdose deaths were collected for 2014-2015 was significantly different than for the years 2016-2023. We opted to drop those years and due to those data sets not including as many well-being measures.

@@

### Findings and Visualizations

@@

``` {r}
# Show deaths per-capita for each state
data2 <- health_ranking |>
  mutate(od_deaths_percapita = combined_overdose_deaths / population)

# All states and overdose death per-capita for years 2016, 2020, 2023
data2 |>
  filter(year %in% c(2016, 2018, 2020, 2023) & !is.na(od_deaths_percapita)) |>
  ggplot(aes(od_deaths_percapita, y = after_stat(count), fill = state)) +
  scale_x_continuous(
    trans = "log10",
    labels = scales::label_number(scale = 1e6, suffix = "M")  # Adjust scale based on your data
  ) +
  geom_density(alpha = 0.3, position = "stack") +
  facet_grid(year ~ .) +
  labs(
    title = "Density Plot of Overdose Deaths Per Capita by State",
    x = "Overdose Deaths Per Capita",
    y = "Count of Overdose Deaths",
    fill = "State"
  )

```

```{r}
#Create individual data sets for Oregon, Washington, and California
or_health_ranking <- health_ranking[health_ranking$state == "Oregon",]
ca_health_ranking <- health_ranking[health_ranking$state == "California",]
wa_health_ranking <- health_ranking[health_ranking$state == "Washington",]

#Now create yearly averages for the states from the healthrank_to_yearly function from Functions.R:
or_yearly <- healthrank_to_yearly(or_health_ranking)
ca_yearly <- healthrank_to_yearly(ca_health_ranking)
wa_yearly <- healthrank_to_yearly(wa_health_ranking)

ggplot()+
  geom_line(data = or_yearly, aes(x = year, y = combined_overdose_deaths, color = "Oregon")) +
  geom_line(data = ca_yearly, aes(x = year, y = combined_overdose_deaths, color = "California")) +
  geom_line(data = wa_yearly, aes(x = year, y = combined_overdose_deaths, color = "Washington")) +
  scale_color_manual(name = "State",
                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
  labs(y = "Overdose Deaths") +
  theme_bw()

#Now do a log transformation to better show percent differences due to population size differences
ggplot()+
  geom_line(data = or_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Oregon")) +
  geom_line(data = ca_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "California")) +
  geom_line(data = wa_yearly, aes(x = year, y = log(combined_overdose_deaths), color = "Washington")) +
  scale_color_manual(name = "State",
                     values = c("Oregon" = "blue", "California" = "red", "Washington" = "green")) +
  theme_bw()
```

@@

```{r Cleveland dot plots, fig.height=5, fig.width=8}

# Create Cleveland dot plots for each state to show changes in county before and after the policy change 
# Arrange the data mortality in desc order and subset into states to see the changes in county by year
health_ranking_OR <- health_ranking |>
  arrange(desc(combined_overdose_mortality_rate)) |>
  filter(state == "Oregon")

health_ranking_WA <- health_ranking |>
  arrange(desc(combined_overdose_mortality_rate)) |>
  filter(state == "Washington")

health_ranking_CA <- health_ranking |>
  arrange(desc(combined_overdose_mortality_rate)) |>
  filter(state == "California")

# Oregon
health_ranking_OR %>%
  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in Oregon",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()

# Washington
health_ranking_WA %>%
  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in Washington",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()
```

```{r CA Cleveland dot plots, fig.height=7, fig.width=8}
# California 
# Save the ggplot to a variable

health_ranking_CA %>%
  filter(year %in% c(2016, 2020, 2023) & !is.na(combined_overdose_mortality_rate)) %>%
  mutate(county = reorder(county, combined_overdose_mortality_rate)) %>%
  ggplot(aes(combined_overdose_mortality_rate, county)) +
  geom_line(aes(group = county)) +
  geom_point(size = 2, aes(color = factor(year))) +
  labs(title = "Overdose Mortality Rate in California",
       x = "Overdose Mortality Rate",  # Add x-axis label
       y = "County") +
  theme_minimal()

```

### Methods

### Results
Decriminalizing drugs in Oregon has slowed the increase in drug overdose deaths.

### Limitations 
It is a challenge to find a single cause for an impact on drug decriminalization and its relationship to drug overdose deaths. Since multiple factors can contribute to drug-related outcomes. There has been some evidence linked to the Covid-19 pandemic contributing to drug use, well-being, and mental health. There is limited generalizability to this study as findings from synthetic control analysis are specific to the context of Oregon drug policies and may not be generalized to other regions or policies. There is also limited data post treatment period (2 years). Additionally, the synthetic Oregon may not be a good fit. This will be further analyzed throughout the semester. As well as continuing to fully understand the statistical basis for synthetic control better.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Exploratory Analysis},
  pdfauthor={James Santarpio and Mardi Elings},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Exploratory Analysis}
\author{James Santarpio and Mardi Elings}
\date{2023-11-30}

\begin{document}
\maketitle

\hypertarget{data-analytics-term-project}{%
\section{Data Analytics Term
Project}\label{data-analytics-term-project}}

\hypertarget{project-desription}{%
\subsection{Project Desription}\label{project-desription}}

\hypertarget{topic}{%
\subsubsection{Topic:}\label{topic}}

The effect of drug decriminalization in Oregon on overdose deaths and
well-being compared to California and Washington.

\hypertarget{why-it-matters}{%
\subsubsection{Why it Matters}\label{why-it-matters}}

The study of drug decriminalization and overdose rates in Oregon has
potential for significant policy and societal relevance. There are many
debates on the benefits and drawbacks that drug use and
decriminalization has on potential for policy implications involving
economics, public health, criminal justice, and community well-being.

\hypertarget{research-question-and-motivation}{%
\subsubsection{Research Question and
Motivation}\label{research-question-and-motivation}}

What is the effect of drug decriminalization on overdose deaths and
well-being? A policy change like Oregon's can have significant
consequences on populations who are most at risk of overdose deaths,
economic impacts on health care costs, and overall well-being.
Investigating the effects with county level data, we used a
difference-in-difference model with fixed effects to assess the impact
of the policy change in Oregon. By comparing the outcomes in Oregon
before and after the intervention to a control group of Washington and
California counties as control groups.

\hypertarget{description-of-data}{%
\subsubsection{Description of Data}\label{description-of-data}}

\hypertarget{overdose-mortality-by-county-in-the-us}{%
\paragraph{Overdose Mortality by County in the
US:}\label{overdose-mortality-by-county-in-the-us}}

The Overdose Mortality by County in the US is from the CDC Center for
Health Statistics for the years 2020-2023. These provisional
county-Level drug Overdose death counts are based on death records and
is received by the NCHS receives the death records from the state vital
registration offices through the Vital Statistics Cooperative Program
(VSCP). The provisional counts are reported by the county were the
decedent resided, not necessarily where they died.

\hypertarget{county-well-being-rankings}{%
\paragraph{County Well-Being
Rankings:}\label{county-well-being-rankings}}

The County Well-Being Rankings are for the years 2014-2023 and is
collected by County Health Rankings \& Roadmaps CHR\&R, a program of the
University of Wisconsin Population Health Institute. The health rankings
include deaths which include overdose deaths, years of potential life
lost, mental health statistics, income, and other health public health
rankings.

\hypertarget{overdose-by-drug}{%
\paragraph{Overdose by Drug:}\label{overdose-by-drug}}

The Overdose by drug dataset contains information on non-fatal overdose
data by state for 2018-2023. Provided by the CDC, DOSE (Drug Overdose
Surveillance and Epidemiology) collects health department reports in
each state.

@@

\hypertarget{how-are-they-related}{%
\paragraph{How are they related?}\label{how-are-they-related}}

The Well-Being Ranks, Overdose Mortality by County contain county fips
codes, county names and year in relation. The Overdose by Drug has a
relationship of total non-fatal drug overdoses in each state.

\hypertarget{data-processing}{%
\subsection{Data Processing}\label{data-processing}}

For the county health rankings data, we originally had years 2014-2023,
but after further inspection we found the way overdose deaths were
collected for 2014-2015 was significantly different than for the years
2016-2023. We opted to drop those years and due to those data sets not
including as many well-being measures.

@@

\hypertarget{findings-and-visualizations}{%
\subsubsection{Findings and
Visualizations}\label{findings-and-visualizations}}

@@

\includegraphics{Draft1Rmd_files/figure-latex/unnamed-chunk-1-1.pdf}

\includegraphics{Draft1Rmd_files/figure-latex/unnamed-chunk-2-1.pdf}
\includegraphics{Draft1Rmd_files/figure-latex/unnamed-chunk-2-2.pdf}

@@

\includegraphics{Draft1Rmd_files/figure-latex/Cleveland dot plots-1.pdf}
\includegraphics{Draft1Rmd_files/figure-latex/Cleveland dot plots-2.pdf}

\includegraphics{Draft1Rmd_files/figure-latex/CA Cleveland dot plots-1.pdf}

\hypertarget{methods}{%
\subsubsection{Methods}\label{methods}}

\hypertarget{results}{%
\subsubsection{Results}\label{results}}

Decriminalizing drugs in Oregon has slowed the increase in drug overdose
deaths.

\hypertarget{limitations}{%
\subsubsection{Limitations}\label{limitations}}

It is a challenge to find a single cause for an impact on drug
decriminalization and its relationship to drug overdose deaths. Since
multiple factors can contribute to drug-related outcomes. There has been
some evidence linked to the Covid-19 pandemic contributing to drug use,
well-being, and mental health. There is limited generalizability to this
study as findings from synthetic control analysis are specific to the
context of Oregon drug policies and may not be generalized to other
regions or policies. There is also limited data post treatment period (2
years). Additionally, the synthetic Oregon may not be a good fit. This
will be further analyzed throughout the semester. As well as continuing
to fully understand the statistical basis for synthetic control better.

\end{document}
